Stochastic Blockmodels for Relational Event Data

### 2012-03-31 docs/submission.tex
Typo in split-merge description.
debugging splitmerge

\begin{abstract}
Observing network data as a stream of events provides an opportunity to understand differences in how nodes behave given the past.  Recent methods for continuous-time network data model the rate of dyadic events conditioned on the observed history of events and covariates about the nodes.  Methods for static networks, such as stochastic blockmodels, often account for node-level heterogeneity by assuming latent groups of individuals have similar tendencies in their group-wise interactions.  We propose to combine these two approaches by modeling the event dynamics within and between clusters of nodes.  The method is illustrated using dyadic interaction data -- e.g. email, Twitter direct messages.  We show parameter estimates from the model reveal heterogeneity in the dynamics among groups of individuals.  The fitted models have better predictive accuracy than both baselines and relational event models without the latent structure.  Our approach illustrates the efficacy of combining a detailed model for local dependencies and a latent variable model for meso-scale dependencies.
\end{abstract}

@@

Network data, however, is often collected as a sequence of events occurring over time.   Recently work leverages  continuous time models from event history analysis  \cite{AalenOddO.2008} to model network-based event data \cite{Butts2008,Brandes2009,Perry2011,Stadtfeld2010,Stadtfeld2011,Opsahl2011,Vu2011,Vu2011a}.  These models allow one to specify how the process depends on the previous history of events.  In this way one may investigate theories about the underlying processes and make predictions about future data conditioned on the past.

The above continuous time network models assume all nodes in the network behave according to identical dynamics.  In many cases this is unrealistic.  For example, in a small organization comprised of several teams, each team may exhibit different patterns for collaborating via email --  individuals in one group may respond more quickly, while another group may preferentially send to highly active individuals.  %, and in many cases emails between one pair of individuals will not affect emailing behavior between a separate pair of individuals.

@@

One can extend the above to marked point processes where each event contains additional information.  In the context of relational events occurring among $N$ nodes, each event in the process contains both a sender $i$ and a recipient $j$ such that  $(i,j) \in \mathcal{R}$, where $\mathcal{R}$ is the \emph{risk set} comprised of all allowed dyads.  We assume each dyad $(i,j) \in \mathcal{R}$ occurs as a Poisson process with intensity $\lambda_{ij}(t)$ that is piecewise constant.  We can then write the likelihood as  
\begin{align}
\mathcal{L}(\mathcal{A}|\theta) &= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{(i,j) \in \mathcal{R}_{i_m,j_m}}\exp\{ - (t_m - \tau_{mij}) \lambda_{ij}(t_m | z_i,z_j) \}
\label{eqn:llk}
\end{align}
\noindent where event $m$ is the dyad $(i_m,j_m)$, $\tau_{mij}$ is the time  of the changepoint for $\lambda_{i,j}$ prior to the $m$th event, and $\mathcal{R}_{i,j}$ is the set of dyads whose intensity changes if $(i,j)$ occurs. \footnote{We further assume 1) all intensity functions change at times $t=0$ and $t=t_M$, and 2) the first event is drawn uniformly from the risk set.}

@@

For a given dyad $(i,j)$ and time $t$ we may also have a vector of statistics $\mathbf{s}(t,i,j|\mathcal{A}_t,\mathbf{z})$ that is a function of the previous event history  $\mathcal{A}_t = \{(t_m,i_m,j_m): t_m \in [0,t)\}$.   In practice we may believe an interaction between one dyad should not affect the rate of an entirely separate dyad.  As shown in Figure \ref{fig:example}, we incorporate this idea by assuming each intensity function $\lambda_{ij}(t)$ may only change following an event involving either $i$ or $j$.  This implies that  $\mathcal{R}_{i,j}$ is the set of dyads involving either $i$ or $j$ and that we will only use  statistics $\mathbf{s}(t,i,j)$ that do not change until an event involving either $i$ or $j$ occurs.

We hope to learn about the dynamics between subsets of nodes.   To facilitate this, we assume node $i$ is assigned a latent cluster $z_i$ and the intensity functions via a log linear model
\begin{align*}
\log \lambda_{ij}(t | \mathcal{A}_t,\mathbf{z}) &= \boldsymbol{\beta}_{z_i,z_j} \mathbf{s}(t,i,j|\mathcal{A}_t) &
z_i &\sim \mbox{CRP}(\alpha) &
\beta_{k,l,p} &\sim \mbox{N}(\mu_p,\sigma_p^2)
\end{align*}
where for each pair of clusters $(z_i,z_j)$ we have a vector of parameters $\boldsymbol{\beta}_{z_i,z_j}$ that corresponds to the vector of statistics $\mathbf{s}(t,i,j|\cdot)$.   Thus, the model of dynamics for the dyad $(i,j)$ has the same parameters as for other dyads occurring between group $z_i$ and $z_j$.  We allow  the blocks to share information by placing a hierarchical prior on the collection of $\boldsymbol{\beta}_{k.l}$.

@@

In Table \ref{tab:stats} we list the statistics we use in our experiments for  $\mathbf{s}(t,i,j)$.  Several statistics use $v_{mij} \in [0,M]$, the index of the last event where $(i,j)$ occurred (n.b. $t_{v_{mij}} = \tau_{mij}$). If $(i,j)$ has never occurred, we set $v_{mij}=0$.  Here we include degree effects to model possible rich-get-richer phenomenon.  For example, a particular node sending often may indicate they will continue to be a sender in the future.  We normalize these counts by the number of events up until a dyad's prior changepoint.\footnote{This seems to avoid ``explosion'' when simulating from the model.  TODO: Explore the theory about these sorts of statistics.  Aalen has a few suggestions.} Other statistics could be of interest for particular substantive questions \cite{Butts2008,Vu2011}.  \footnote{Describe the form of $f()$.  Right now it is $f(x) = \log \frac{x+1}{m + N(N-1)}$.}

@@

We Gibbs sample the latent class assignments $\mathbf{z}$ from the conditional distribution
\begin{align*}
p(z_i | z_{-i},\mathcal{A},\alpha,\boldsymbol{\beta}) &\propto p(z_i | z_{-i},\alpha) p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta},\mbox{node} \ i \ \mbox{involved}) \\
p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta},\mbox{node} \ r \ \mbox{involved}) &= \prod_{a \in \mathcal{U}_r} \lambda_{i_a,j_a}(t_a|\cdot)
\prod_{(i,j) \in \mathcal{R}_{i_a,j_a}} \exp \{ -(t_a - \tau_{aij}) \lambda_{ij}(t_a|\cdot)\}
\end{align*}
where $\mathcal{U}_r = \{a: r \in \{i_a,j_a\}, a \in [1,M]\}$ is the set of events where node $r$ is involved.  Under a CRP($\alpha$) prior, we have $p(z_i = k | z_{-i},\alpha) = n_k $ and $p(z_i = \mbox{new} |  z_{-i},\alpha) = \alpha$ where $n_k$ is the number of nodes assigned to cluster $k$. 
% \begin{align*}
% p(z_i = k | z_{-i},\alpha) &= n_k & p(z_i = \mbox{new} |  z_{-i},\alpha) &= \alpha
% \end{align*}

 By limiting the changepoints to times when either node is involved, computing the likelihood $p(A|z,\beta,\mbox{node} \ r \ \mbox{involved})$ is $O(|\mathcal{U}_r| \cdot P \cdot 4N)$.  The factor of 4 is due to the number of dyads involving a single node.  We precompute $\tau_{m,i,j}$ and $\mathbf{s}(t_m,i,j)$ for all $m,i,j$.

@@

For each block $(k,l)$ we need to sample the vector of parameters $\boldsymbol{\beta}_{k,l}$ from its posterior
\begin{align*}
p(\boldsymbol{\beta}_{k,l} | \mathcal{A}_t, \textbf{z}, \mu, \sigma) &\propto p(\boldsymbol{\beta}_{k,l} | \mu, \sigma) p( \mathcal{A}| \textbf{z}, \boldsymbol{\beta}, \mbox{cluster} \ k \ \mbox{or} \ l \ \mbox{involved}) \\  
p(\boldsymbol{\beta}_{k,l} | \mu, \sigma) &= \prod_{p=1}^Pp(\beta_{k,l,p}|\mu_p,\sigma_p^2)\\
p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta}, \mbox{cluster} \ k \ \mbox{or} \ l \ \mbox{involved}) &= \prod_{a \in \mathcal{U}_r} \lambda_{i_a,j_a}(t_a|\cdot)
\prod_{(i,j) \in \mathcal{R}_{i_a,j_a}} \exp \{ -(t_a - \tau_{aij}) \lambda_{ij}(t_a|\cdot)\}
\end{align*}
where $\mathcal{V}_{k,l} = \{a: z_{i_a} \in \{k,l\} \ \mbox{or} \ z_{j_a} \in \{k,l\}, a \in [1,M]\}$ is the set of events $a$ where one of the nodes is either in block $k$ or $l$.  \footnote{TODO: If we end up using HMC, mention step size and step length used.}
This distribution is sampled using slice sampling.\footnote{We may also use HMC for this.}

@@

The first two terms represent the prior. For the proposed merge step, note $q(\boldsymbol{\phi}^{split} \rightarrow \boldsymbol{\phi}^{merge}) = .5^{P(2K-3)}$ where $K$ is the number of blocks before the move.  We can compute $q(\boldsymbol{\phi}^{merge} \rightarrow \boldsymbol{\phi}^{split})$ directly as a product of Normal densities.  For the block assignments, $q(\mathbf{z}^{split} \rightarrow \mathbf{z}^{merge})=1$ since  there is only one way of combining two clusters.  $ \frac{p(\mathbf{z}^{split})}{p(\mathbf{z}^{merge})} = \frac{(n_{z_i}^{split} - 1)!(n_{z_j}^{split} - 1)!}{(n_{z_i}^{merge}-1)!}$.  We have avoided the dimensionality of the model during the MH move by creating an empty cluster when merging and filling a (newly created) empty cluster when splitting.  One can return to the previous state with positive probability...

@@

Though the dyad counts for the synthetic data set suggest a stochastic blockmodel (as seen in Figure  \ref{fig:syncounts}),  the center plot shows each block has empirical differences in their dynamics.  Intensities for reciprocal actions among nodes in block 1 are $e^3$ times greater, intensities for turn-taking actions among nodes in group 2 are $e^2$ times greater, and intensities for dyadic interactions between the two groups have a multiplicative effect of $e^{-1}$ and thus occur less often.  Fitting the model with $K=2$ converges to the true loglikelihood (see Table \ref{tab:results}) and the posterior credible intervals of the parameters cover the true parameter values.

@@

Figure \ref{fig:posteriorparams} shows examples of model estimates from data sets of dyadic interactions. Panels contain posterior credible intervals for parameters pertaining to dyads belonging to a particular block.  For example, events among nodes in cluster 1 have an increased propensity for reciprocal events -- after node A sends to node B, an event from B to A has an intensity that is $e^4$ larger, all else held constant.  Interestingly, degree effects seem to vary by block: in block (2,2) active senders are more likely to be a sender whereas in block (1,1) the number of previous occurrences for a given dyad has a larger multiplicative effect.

@@

>>>>

\begin{abstract}
Observing network data as a stream of relational events provides an
opportunity to understand differences in how nodes interact given the
past.  For continuous-time network data, recent methods model the rate of dyadic events conditioned on the observed history of events and covariates about the nodes.  For static networks, methods such as stochastic blockmodels often account for node-level heterogeneity by assuming latent groups of individuals have similar tendencies in their group-wise interactions.  We propose to combine these two approaches by modeling the event dynamics within and between clusters of nodes.  The method is illustrated using dyadic interaction data such as email and Twitter direct messages.  Parameter estimates from the model clearly reveal heterogeneity in the dynamics among groups of individuals.  The fitted models have better predictive accuracy than baselines and relational event models without the latent structure.  Our approach illustrates the efficacy of combining a detailed model for local dependencies and a latent variable model for meso-scale dependencies.
\end{abstract}

@@

Network data, however, is often collected as a sequence of events occurring over time.   Recent work leverages  continuous-time models from event history analysis  \cite{AalenOddO.2008} to model network-based event data \cite{Butts2008,Brandes2009,Perry2011,Stadtfeld2010,Stadtfeld2011,Opsahl2011,Vu2011,Vu2011a}.  These models allow one to specify how the process depends on the previous history of events.  In this way one may investigate theories about the underlying processes and make predictions about future data conditioned on the past.

The above continuous time network models assume all nodes in the network behave according to identical dynamics.  In many cases this is unrealistic.  For example, in a small organization comprised of several teams, each team may uncover different patterns for collaborating via email --  individuals in one group may respond more quickly, while another group may preferentially send to highly active individuals.  %, and in many cases emails between one pair of individuals will not affect emailing behavior between a separate pair of individuals.

@@

One can extend the above to marked point processes where each event contains additional information.  In the context of relational events occurring among $N$ nodes, each event in the process contains both a sender $i$ and a recipient $j$ such that  $(i,j) \in \mathcal{R}$, where $\mathcal{R}$ is the \emph{risk set} comprised of all allowed dyads.  We assume each dyad $(i,j) \in \mathcal{R}$ occurs as a Poisson process with intensity $\lambda_{ij}(t)$ that is piecewise constant.  We can then write the likelihood as  
\begin{align}
\mathcal{L}(\mathcal{A}|\theta) &= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{(i,j) \in \mathcal{R}_{i_m,j_m}}\exp\{ - (t_m - \tau_{mij}) \lambda_{ij}(t_m | z_i,z_j) \}
\label{eqn:llk}
\end{align}
\noindent where event $m$ is the dyad $(i_m,j_m)$, $\tau_{mij}$ is the time  of the changepoint for $\lambda_{i,j}$ prior to the $m$th event, and $\mathcal{R}_{i,j}$ is the set of dyads whose intensity changes if $(i,j)$ occurs  \cite{Butts2008}. \footnote{We further assume 1) all intensity functions change at times $t=0$ and $t=t_M$, and 2) the first event is drawn uniformly from the risk set.}

@@

For a given dyad $(i,j)$ and time $t$ we may also have a vector of statistics $\mathbf{s}(t,i,j|\mathcal{A}_t,\mathbf{z})$ that is a function of the previous event history  $\mathcal{A}_t = \{(t_m,i_m,j_m): t_m \in [0,t)\}$.   In practice we may believe an interaction between one dyad should not affect the rate of an entirely separate dyad.  As shown in Figure \ref{fig:example}, we incorporate this idea by assuming each intensity function $\lambda_{ij}(t)$ can only change following an event involving either $i$ or $j$.  This implies that  $\mathcal{R}_{i,j}$ is the set of dyads involving either $i$ or $j$ and that we will only use  statistics $\mathbf{s}(t,i,j)$ that do not change until an event involving either $i$ or $j$ occurs.

We hope to learn about the dynamics between subsets of nodes.   To facilitate this, we assume each node $i$ is assigned a latent cluster $z_i$ and use a log linear model for the intensity functions
\begin{align*}
\log \lambda_{ij}(t | \mathcal{A}_t,\mathbf{z}) &= \boldsymbol{\beta}_{z_i,z_j} \mathbf{s}(t,i,j|\mathcal{A}_t) &
z_i &\sim \mbox{CRP}(\alpha) &
\boldsymbol{\beta}_{k,l} &\sim \mbox{N}_p(\boldsymbol{\mu},\boldsymbol{\sigma}^2I)
\end{align*}
where for each pair of clusters $(z_i,z_j)$ we have a vector of parameters $\boldsymbol{\beta}_{z_i,z_j}$ that corresponds to the vector of statistics $\mathbf{s}(t,i,j|\cdot)$.   Thus, the model of dynamics for the dyad $(i,j)$ has the same parameters as for other dyads occurring between group $z_i$ and $z_j$.  We allow  the blocks to share information by placing a hierarchical prior on the collection of $\boldsymbol{\beta}_{k.l}$ where $\mu_p \sim N(0,1)$ and $\sigma_p \sim \mbox{Gamma}(3,1)$ (TODO).

@@

In Table \ref{tab:stats} we list the statistics we use in our experiments for  $\mathbf{s}(t,i,j)$.  Several statistics use $v_{mij} \in [0,M]$, the index of the last event where $(i,j)$ occurred (n.b. $t_{v_{mij}} = \tau_{mij}$). If $(i,j)$ has never occurred, we set $v_{mij}=0$.  We include degree effects to model possible rich-get-richer phenomenon.  For example, a particular node sending often may indicate they will continue to be a sender in the future.  We normalize these counts by the number of events up until a dyad's prior changepoint.\footnote{This seems to avoid ``explosion'' when simulating from the model.  TODO: Explore the theory about these sorts of statistics.  Aalen has a few suggestions.} Other statistics could be of interest for particular substantive questions \cite{Butts2008,Vu2011}.  \footnote{Describe the form of $f()$.  Right now it is $f(x) = \log \frac{x+1}{m + N(N-1)}$.}

@@

We use Gibbs sampling to sample the latent class assignments $\mathbf{z}$ from the conditional distribution
\begin{align*}
p(z_i | \mathbf{z}_{-i},\mathcal{A},\alpha,\boldsymbol{\beta}) &\propto p(z_i | \mathbf{z}_{-i},\alpha) p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta},\mbox{node} \ i \ \mbox{involved}) \\
p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta},\mbox{node} \ r \ \mbox{involved}) &= \prod_{a \in \mathcal{U}_r} \lambda_{i_a,j_a}(t_a|\cdot)
\prod_{(i,j) \in \mathcal{R}_{i_a,j_a}} \exp \{ -(t_a - \tau_{aij}) \lambda_{ij}(t_a|\cdot)\}
\end{align*}
where $\mathcal{U}_r = \{a: r \in \{i_a,j_a\}, a \in [1,M]\}$ is the set of events where node $r$ is involved.  Under a CRP($\alpha$) prior, we have $p(z_i = k | z_{-i},\alpha) = n_k $ and $p(z_i = \mbox{new} |  z_{-i},\alpha) = \alpha$ where $n_k$ is the number of nodes assigned to cluster $k$. 
% \begin{align*}
% p(z_i = k | z_{-i},\alpha) &= n_k & p(z_i = \mbox{new} |  z_{-i},\alpha) &= \alpha
% \end{align*}

 By limiting the changepoints to times when either node is involved, computing the likelihood $p(A|z,\beta,\mbox{node} \ r \ \mbox{involved})$ is $O(|\mathcal{U}_r| \cdot P \cdot N)$.  We precompute $\tau_{m,i,j}$ and $\mathbf{s}(t_m,i,j)$ for all $m,i,j$.

@@

For each block $(k,l)$ we need to sample the vector of parameters $\boldsymbol{\beta}_{k,l}$ from its posterior
\begin{align*}
p(\boldsymbol{\beta}_{k,l} | \mathcal{A}_t, \textbf{z}, \boldsymbol{\mu}, \boldsymbol{\sigma}) &\propto p(\boldsymbol{\beta}_{k,l} | \boldsymbol{\mu}, \boldsymbol{\sigma}) p( \mathcal{A}| \textbf{z}, \boldsymbol{\beta}, \mbox{cluster} \ k \ \mbox{or} \ l \ \mbox{involved}) \\  
p(\boldsymbol{\beta}_{k,l} | \mu, \sigma) &= \prod_{p=1}^Pp(\beta_{k,l,p}|\mu_p,\sigma_p^2)\\
p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta}, \mbox{cluster} \ k \ \mbox{or} \ l \ \mbox{involved}) &= \prod_{a \in \mathcal{U}_r} \lambda_{i_a,j_a}(t_a|\cdot)
\prod_{(i,j) \in \mathcal{R}_{i_a,j_a}} \exp \{ -(t_a - \tau_{aij}) \lambda_{ij}(t_a|\cdot)\}
\end{align*}
where $\mathcal{V}_{k,l} = \{a: z_{i_a} \in \{k,l\} \ \mbox{or} \ z_{j_a} \in \{k,l\}, a \in [1,M]\}$ is the set of events $a$ where one of the nodes is either in block $k$ or $l$.  \footnote{TODO: If we end up using HMC, mention step size and step length used.}
This distribution is sampled using slice sampling.\footnote{We may also use HMC for this.}

@@

The first two terms represent the prior. For the proposed merge step, note $q(\boldsymbol{\phi}^{split} \rightarrow \boldsymbol{\phi}^{merge}) = .5^{P(2K-3)}$ where $K$ is the number of blocks before the move.  We can compute $q(\boldsymbol{\phi}^{merge} \rightarrow \boldsymbol{\phi}^{split})$ directly as a product of Normal densities.  For the block assignments, $q(\mathbf{z}^{split} \rightarrow \mathbf{z}^{merge})=1$ since  there is only one way of combining two clusters.  $ \frac{p(\mathbf{z}^{split})}{p(\mathbf{z}^{merge})} = \frac{(n_{z_i}^{split} - 1)!(n_{z_j}^{split} - 1)!}{(n_{z_i}^{merge}-1)!}$.  We have avoided the dimensionality of the model during the MH move by creating an empty cluster when merging and filling a (newly created) empty cluster when splitting.  One can return to the previous state with positive probability.

@@

Though the dyad counts for the synthetic data set suggest a stochastic blockmodel (as seen in Figure  \ref{fig:syncounts}),  the center plot shows each block has empirical differences in their dynamics.  Intensities for reciprocal actions among nodes in block 1 are $e^3$ times greater, intensities for turn-taking actions among nodes in group 2 are $e^2$ times greater, and intensities for dyadic interactions between the two groups have a multiplicative effect of $e^{-1}$ and thus occur less often.  Fitting the model with $K=2$ makes the same predictions as the true model (see Table \ref{tab:results}) and the posterior credible intervals of the parameters cover the true parameter values.

@@

Figure \ref{fig:posteriorparams} shows examples of model estimates from data sets of dyadic interactions for the Eckmann data. Panels contain posterior credible intervals for parameters pertaining to dyads belonging to a particular block.  For example, events among nodes in cluster 1 have an increased propensity for reciprocal events -- after node A sends to node B, an event from B to A has an intensity that is $e^4$ larger, all else held constant.  Interestingly, degree effects seem to vary by block: in block (2,2) active senders are more likely to be a sender whereas in block (1,1) the number of previous occurrences for a given dyad has a larger multiplicative effect.

@@

\appendix{Split-merge updates}

The first two terms represent the prior. For the proposed merge step, note $q(\boldsymbol{\phi}^{split} \rightarrow \boldsymbol{\phi}^{merge}) = .5^{P(2K-3)}$ where $K$ is the number of blocks before the move.  We can compute $q(\boldsymbol{\phi}^{merge} \rightarrow \boldsymbol{\phi}^{split})$ directly as a product of Normal densities.  For the block assignments, $q(\mathbf{z}^{split} \rightarrow \mathbf{z}^{merge})=1$ since  there is only one way of combining two clusters.  $ \frac{p(\mathbf{z}^{split})}{p(\mathbf{z}^{merge})} = \frac{(n_{z_i}^{split} - 1)!(n_{z_j}^{split} - 1)!}{(n_{z_i}^{merge}-1)!}$.  We have avoided the dimensionality of the model during the MH move by creating an empty cluster when merging and filling a (newly created) empty cluster when splitting.  One can return to the previous state with positive probability.
